Visualising CHD Pathway Inequalities by ethnicity

### 2024-07-24 final_report.qmd
Add data tables, download buttons and conclusion

---
title: "Visualising CHD Pathway Inequalities by ethnicity - DRAFT"
lang: en-GB
author: "Jacqueline Grout"
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
format:
  html:
    self-contained: true
    grid:
      sidebar-width: 200px
      body-width: 950px
      margin-width: 150px
      gutter-width: 1.5rem
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Contents
    toc-depth: 3
editor: visual
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
editor_options: 
  chunk_output_type: console

@@

library(targets)
library(gt)
library(tidyverse)
library(grid)
library(gridExtra)

@@

@@

@@

```{r}
#1
region_cluster_charts <- tar_read(region_cluster_charts)
region_cluster_charts[2]
```

@@

```{r}
#2
region_cluster_charts[7]
```

@@

```{r}
#3
region_cluster_charts[6]
```

@@

```{r}
#4
region_cluster_charts[3]
```

@@

```{r}
#5
region_cluster_charts[1]
```

@@

```{r}
#6
region_cluster_charts[5]
```

@@

```{r}
#7
region_cluster_charts[4]
```
:::

@@

To be added as part of finalisation of the report.

@@

```{r}
icb_charts <- tar_read(icb_charts)
icb_charts[1]
icb_charts[2]
icb_charts[3]
icb_charts[4]
icb_charts[5]
icb_charts[6]
icb_charts[7]
icb_charts[8]
icb_charts[9]
icb_charts[10]
icb_charts[11]
icb_charts[12]
icb_charts[13]
icb_charts[14]
icb_charts[15]
icb_charts[16]
icb_charts[17]
icb_charts[18]
icb_charts[19]
icb_charts[20]
icb_charts[21]
icb_charts[22]
icb_charts[23]
icb_charts[24]
icb_charts[25]
icb_charts[26]
icb_charts[27]
icb_charts[28]
icb_charts[23]
icb_charts[29]
icb_charts[30]
icb_charts[31]
icb_charts[32]
icb_charts[33]
icb_charts[34]
icb_charts[35]
icb_charts[36]
icb_charts[37]
icb_charts[38]
icb_charts[39]
icb_charts[40]
icb_charts[41]
icb_charts[42]
```

>>>>

---
title: "Visualising CHD Pathway Inequalities by ethnicity"
lang: en-GB
author: "Jacqueline Grout"
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
format:
  html:
    self-contained: true
    grid:
      sidebar-width: 200px
      body-width: 950px
      margin-width: 150px
      gutter-width: 1.5rem
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Contents
    toc-depth: 3
editor: visual
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
editor_options: 
  chunk_output_type: console

@@

library(targets)
library(gt)
library(tidyverse)
library(grid)
library(gridExtra)
library(downloadthis)

@@

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
## Activity rates data

```{r}
rate_chart_data <- tar_read(rate_chart_data) |> as_tibble()

rate_chart_data |> 
  dplyr::rename(cluster=cluster2)|>
  dplyr::select(-name) |>
  gt::gt() |>
  gt::cols_move(columns = c(rate,lower_ci,upper_ci, global_rate),
                after = metric_name) |>
  gt::fmt_number(columns=c(rate,lower_ci,upper_ci, global_rate),
                 decimals=3) |>
  gt::cols_label(cluster =gt::md("**Cluster**"),
                 pathway = gt::md("**Pathway**"),
                 metric_name = gt::md("**Metric**"),
                 rate= gt::md("**Rate**"),
                 lower_ci = gt::md("**Lower CI**"),
                 upper_ci = gt::md("**Upper CI**"),
                 global_rate = gt::md("**Global Rate**")) |>
  gt::tab_row_group(
    label = gt::md("**Full outcomes**"),
    rows = pathway == "Full outcomes"
  ) |>
  gt::tab_row_group(
    label = gt::md("**Intermediate outcome**"),
    rows = pathway == "Intermediate outcome"
  ) |>
  gt::tab_row_group(
    label = gt::md("**Tertiary prevention**"),
    rows = pathway == "Tertiary prevention"
  ) |>
  gt::tab_row_group(
    label = gt::md("**Secondary prevention**"),
    rows = pathway == "Secondary prevention"
  ) |>
  gt::tab_row_group(
    label = gt::md("**Disease identification**"),
    rows = pathway == "Disease identification"
  ) |>
  gt::tab_row_group(
    label = gt::md("**Primary prevention**"),
    rows = pathway == "Primary prevention"
  ) |>
  gt::tab_row_group(
    label = gt::md("**Risk factor identification**"),
    rows = pathway == "Risk factor identification"
  ) |>
  gt::tab_row_group(
    label = gt::md("**Risk factors**"),
    rows = pathway == "Risk factors"
  ) 
```
:::

```{r}
rate_chart_data %>%
  download_this(
    output_name = "rate_data",
    output_extension = ".xlsx",
    button_label = "Download activity rate data",
    button_type = "default",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```

@@

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
## Index of Disparity data

```{r}
rate_chart_data <- tar_read(rate_chart_data) |> as_tibble()
iod_with_ci <- tar_read(iod_with_ci) |> as_tibble()

metric_names <- rate_chart_data |> 
  select(pathway,metric_name,name) |> 
  unique() |> 
  rename(metric_name_full=metric_name, metric_name=name)

disparity_data <- iod_with_ci |> left_join(metric_names)

disparity_data |> 
  select(-metric_name) |>
  gt() |>
  cols_move(columns = c(iod,lower_ci,upper_ci),
            after = metric_name_full) |>
  tab_row_group(
    label = md("**Full outcomes**"),
    rows = pathway == "Full outcomes"
  ) |>
  tab_row_group(
    label = md("**Intermediate outcome**"),
    rows = pathway == "Intermediate outcome"
  ) |>
  tab_row_group(
    label = md("**Tertiary prevention**"),
    rows = pathway == "Tertiary prevention"
  ) |>
  tab_row_group(
    label = md("**Secondary prevention**"),
    rows = pathway == "Secondary prevention"
  ) |>
  tab_row_group(
    label = md("**Disease identification**"),
    rows = pathway == "Disease identification"
  ) |>
  tab_row_group(
    label = md("**Primary prevention**"),
    rows = pathway == "Primary prevention"
  ) |>
  tab_row_group(
    label = md("**Risk factor identification**"),
    rows = pathway == "Risk factor identification"
  ) |>
  tab_row_group(
    label = md("**Risk factors**"),
    rows = pathway == "Risk factors"
  ) |>

  fmt_number(columns=c(iod,lower_ci,upper_ci),
             decimals=3) |>
  cols_label(pathway = md("**Pathway**"),
             metric_name_full = md("**Metric**"),
             iod= md("**IOD**"),
             lower_ci = md("**Lower CI**"),
             upper_ci = md("**Upper CI**"))
```
:::

```{r}
disparity_data %>%
  download_this(
    output_name = "iod_data",
    output_extension = ".xlsx",
    button_label = "Download iod data",
    button_type = "default",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```

@@

```{r}
#1
region_cluster_charts <- tar_read(region_cluster_charts)
region_cluster_charts[[2]]
```

@@

```{r}
#2
region_cluster_charts[[7]]
```

@@

```{r}
#3
region_cluster_charts[[6]]
```

@@

```{r}
#4
region_cluster_charts[[3]]
```

@@

```{r}
#5
region_cluster_charts[[1]]
```

@@

```{r}
#6
region_cluster_charts[[5]]
```

@@

```{r}
#7
region_cluster_charts[[4]]
```
:::

@@

In conclusion, there are three main aspects to this analysis on which to reflect: methods, findings and next steps.

Assessing inequalities across ethnic groups was challenging on a number of levels, it was both experimental and complex. Ethnicity can’t be expressed as a set of ordered groups and there is no meaningful way to numerically aggregate into a single variable. Consequently, relative index of inequality can’t be used.

This analysis attempted to overcome these challenges using novel methods, creating 5 clusters of GP practices using K-Medoids clustering on ethnicity %, followed by use of the relative index of disparity to indicate the extent to which the rate of an activity or event on the CHD pathway varied across the clusters.

Further challenges arose during the analysis, which are detailed in the appendix, however this analysis has made progress in the methods used and they are useful in presenting a visualisation of the variation in disparity of equity across the CHD pathway.

The results of the analysis should be viewed as tentative rather than actionable, adding to a fuller picture alongside other work and thinking on inequalities and CHD.

Early findings show that variation clearly exists across the pathway with some metrics showing less than 1% disparity (blood pressure checks within 5 years for 45+), whilst others have as much as 17% disparity (flu vaccinations for 65+).

The analysis found that risk factors varied little across clusters, although diabetes rates were identified more in the most diverse cluster and depression was identified more in the least diverse cluster. Moving through the pathway, CVD patients or at risk patients had higher rates of lipid lowering therapies in the least diverse clusters. Rates of diagnosis (on the CHD register, CT angiography, and electrocardiography) were all lower in the more diverse clusters.

Moving across the pathway towards treatment and care, inequity showed up more between clusters in some of the secondary and tertiary prevention metrics in particular CHD patients receiving Asprin, APT or ACT (12.5%), starting cardiac rehabilitation(11%) and Elective CABG (10%).

Mortality (CHD deaths and premature CHD deaths, including in hospital) was higher in the least diverse cluster, which may reflect age and sex profiles of the population (and which is examined and discussed in further detail in the appendix).

A lot of risk factors had similar rates to those found when examining the pathway through a socio-economic lens but a lot of diagnostics such as being on the CHD register or getting a CT Angiography were lower in the more diverse clusters. There are also parallels with the previous work examining socio-economic inequalities along the pathway for secondary and tertiary prevention metrics with many of these metrics such as Elective PCI and CABG and receipt of Asprin etc also showing that inequity favoured the least deprived.

Many of these findings give rise to further questions and therefore there are a number of further routes for analysis and possible next steps that could be taken to extend and follow up this project.

Whilst this project includes some regional and ICB level presentations, in some cases the majority of GP practices in a region or ICB have the same cluster. The analysis could be recreated at a regional level, whereby new unique clusters could be assigned separately for a region, potentially giving rise to more nuanced clusters from which local disparities could be more easily identified.

The methods used in this project could be transferred to other pathways and services, re-using the same clusters and then selecting different sets of metrics for example related to COPD, Cancer, Mental Health etc.

The code for this project is available on GitHub and as such this enables the work to be reused and extended more easily.

Many of the metrics used in this analysis were only available at a GP practice level, necessitating some of the approaches taken. Where record level data is available this would allow for more detailed exploration and understanding of some of these initial findings. Of particular interest would be further exploration of the findings surrounding mortality rates and the inter-relationships between different factors of inequality such as ethnicity and deprivation, as well as better understanding at an individual ethnicity level.

@@

```{r}
#| output: "asis"
icb_charts <- tar_read(icb_charts)

purrr::iwalk(icb_charts, \(plot, name) {
  name <- str_extract(name, "(?<=^NHS ).*(?= Integrated Care Board$)")
  cat("\n\n#####", name, "\n\n")
  print(plot)
})
```

all_icb_data <- purrr::map(icb_charts, "data") |> dplyr::bind_rows(.id = "icb")

all_icb_data %>%
  download_this(
    output_name = "iod_icb_data",
    output_extension = ".xlsx",
    button_label = "Download ICB IoD data",
    button_type = "default",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```
